---
title: "Template"
author: "Dr. Islam"
date: "2023-04-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
X<-iris
head(X)
summary(X$Sepal.Length)
```

Q1. 
Is there any significant difference in mean of sepal length and petal length at 5% level of significance?

Hypothesis:

$H_a:\mu_1=\mu_2$ vs $H_a:\mu_1 \ne \mu_2$

The test statistic is $$t=\frac{\bar{x}_1-\bar{x_2}-0}{\sqrt{\frac{s_1^2}{n_1}+\frac{s_2^2}{n_2}}}$$

```{r}
boxplot(X$Sepal.Length, X$Petal.Length, horizontal = T)
t.test(X$Sepal.Length, X$Petal.Length)
```
With the p-value of less than 0.05, we have strong evidence to claim that the mean length of sepal and mean length of petal are significantly different. 






$$t=\frac{\bar{x_1}-\bar{x_2}}{\sqrt{\frac{s_1^2}{n_1}+\frac{s_2^2}{n2}}}$$

$H_0: \mu_1=\mu_2$ vs $H_a: \mu_1 \ne \mu_2$

$H_0: \mu_1=\mu_2$ vs $H_a: \mu_1 \gt \mu_2$

$H_0: \mu_1=\mu_2$ vs $H_a: \mu_1 \lt \mu_2$


$z=\frac{\hat{p_1}-\hat{p_2}}{ \sqrt{\hat{p}(1-\hat{p})(1/n_1+1/n_2)}}$

$$y=\alpha+\beta x+\epsilon$$

$H_0: \beta=0$ vs $H_a:\beta \ne 0$


